A subjective Bayesian framework for synthesizing deep uncertainties in climate risk management

### 2022-06-17 papers/paper/preprint.tex
address klaus handwritten edits

% Figures
\usepackage[demo]{graphicx}
\usepackage[list=true]{subcaption}
\graphicspath{{../../plots/}{../../tikz/}{../../img/}}
\usepackage[section]{placeins} % require floats to appear in the section they are defined

@@

% ACRONYMS
\usepackage[acronym, nopostdot, nonumberlist, shortcuts, numberedsection, nogroupskip,]{glossaries}
\newacronym{bfe}{BFE}{base flood elevation}
\newacronym{cmip}{CMIP}{the Coupled Model Intercomparison Project}
\newacronym{dmdu}{DMDU}{decision making under deep uncertainty}
\newacronym{fema}{FEMA}{the Federal Emergency Management Agency}
\newacronym{gev}{GEV}{generalized extreme value}
\newacronym{iid}{IID}{independent and identically distributed}
\newacronym{ipcc}{IPCC}{International Panel on Climate Change}
\newacronym{msl}{MSL}{mean relative sea level}
\newacronym{noaa}{NOAA}{the National Oceanic and Atmospheric Administration}
\newacronym{pdf}{PDF}{probability density function}
\newacronym{rcp}{RCP}{representative concentration pathway}
\newacronym{slr}{SLR}{sea level rise}
\newacronym{ssp}{SSP}{shared socio-economic pathway}
\newacronym[]{usace}{USACE}{United States Army Corps of Engineers}
\newacronym[]{usgs}{USGS}{United States Geological Survey}
\newacronym[plural=SOWs,firstplural=states of the world (SOWs),descriptionplural=states of the world]{sow}{SOW}{state of the world}
\makeglossaries % required to make the list of acronyms

@@

Currently used methods for estimating these \glspl{pdf} emphasize data-driven, nominally objective methods that can be applied consistently across locations.
For example, \gls{usgs} Bulletin 17C specifies procedures for estimating flood frequency and \gls{noaa} Atlas 14 provides estimates of the intensity, duration, and frequency of extreme rainfall.
Among several statistical assumptions these analyses make that have been recently called into question is stationarity \citep[the assumption that past and future hazard come from the same \gls{pdf}; see][for a review]{milly_stationarity:2008,merz_review:2014}.
For example, clear trends in extreme rainfall are apparent across much of southeastern Texas \citep{fagnant_spatiotemporal:2020,nielsen-gammon_txrainfall:2020}, and trends in many other hazards are consistent with observations \citep[see][for a comprehensive summary]{ipcc_impacts:2022}.
While some methods have been proposed for incorporating trends into these analyses \citep[see][for a review]{Salas:2018ge}, these assume specific forms of nonstationarity which may not fully represent physical processes or true levels of uncertainty \citep{DossGollin:2019,Montanari:2014hl,Serinaldi:2015bq}.
Ultimately, the difficulty of incorporating nonstationarity into existing frameworks has led to continued reliance on stationarity despite recognition of the associated limitations (see \cite{bulletin17c:2019}, p.~2, or \cite{atlas14_texas:2018} p.~A.4-42).

@@

Although these methods for \gls{dmdu} have proven valuable in a wide range of settings, they still require necessarily subjective assessments about the likelihood of future conditions.
Even exploratory models require the analyst to choose which uncertainties are considered and how they are sampled.
When analyses use metrics that integrate performance over many possible futures, whether to compute sensitivities, estimate expectations, or compute robustness metrics, they necessarily make assumptions about the likelihood of different futures.
For example, uniform priors are often justified by deferring to Laplace's principle of insufficient reason, but even this choice makes strong assumptions that are sensitive to factors such as parameterization \citep[p.54]{gelman_bda3:2014}.
While one interpretation of deep uncertainty is that that specifying a joint \gls{pdf} over inputs is inappropriate, assumptions about the ranges and independence of parameters to sample are just as subjective as the choice of probability distribution \citep{schneider_scenarios:2002,quinn_exploratory:2020} and, indeed, can be interpreted  as a specific choice of probability distribution (we revisit this point in \cref{sec:analysis-condition}).
This motivates the development of decision analytic frameworks that draw from the strengths of \gls{dmdu} methods such as exploratory modeling, vulnerability assessment, robustness checks, and iterative stakeholder critique, but that embrace the reality that assumptions about the future are inescapable.

@@

We are motivated by parallels between decision making under deep uncertainty and the statistical problem of model selection.
\citet{oreskes_verification:1994} argues that because natural systems are never closed and model results are never unique, validation and verification of models representing these systems is necessarily qualitative and subjective.
In the literature on decision making under deep uncertainty, this has led to recognition of the need to develop strategies that are robust to model imperfections.
This challenge is described in the Bayesian literature on model selection, that is the problem of using competing statistical models to draw inferences, particularly in the ``$\mathcal{M}$-open'' case where there is no true model to identify and all models are imperfect approximations of reality.
Just as the literature on decision making under deep uncertainty contrasts with classical methods for decision analysis that identify an objectively ``best'' decision (or decision set) under a ``true'' description of uncertaintty, the literature on Bayesian model selection in the $\mathcal{M}$-open case contrasts with classical methods that seek to asymptotically identify a ``true'' model from a candidate space.
In particular, model selection in the $\mathcal{M}$-open case emphasizes a subjective view, both of probability itself and of the modeling process, for which probability offers a self-consistent language with which to reason about the unknown rather than a statement of objective truth (see \cite{gelman_philosophy:2013} or \cite{ramsey_probability:2016} for a philosophical discussion, \cite{jaynes_probability:2003} for a detailed reference, or \cite{Piironen:2017eh} for a technical discussion of methods for model selection).
From a practical perspective, emphasis is placed on iteratively building models, simulating the consequences of those models, and subsequently critiquing them \citep{gelman_workflow:2020}.

In this paper we offer a first conceptual step towards bridging the fields of \gls{dmdu} and Bayesian model building.
We consider the didactic case study of whether to elevate a hypothetical house, and if so how high, as a specific example of a decision problem subject to both shallow (storm surge) and deep (\gls{slr}) uncertainties.
Prior studies have found that floodproofing and building-scale vulnerability reduction measures, including house elevation, can effectively reduce local flood damages in many contexts \citep{demoel_reducing:2014,deruig_building:2020,kreibich_building:2005,slotter_floodproofing:2020,rozer_coping:2016,mobley_mitigation:2020,aerts_cost:2018}, and both local building codes \citep{asce_7-10:2013,bruneau_multihazard:2017,asce_24-05:2006} and federal policy \citep{FEMA_p-55:2011} require elevation in some cases.
Guidance for homeowners, notably from \gls{fema}, recommends elevating to the \gls{bfe} (typically the \SI{100}{year} flood) plus a freeboard \citep{fema_retrofitting:2014,asce_24-14:2015,fema_retrofitting:2014} but recent research has demonstrated that neglecting uncertainty in the cost-benefit analysis can lead to poor decisions \citep{zarekarizi_suboptimal:2020}.
Focusing on deep uncertainty in \gls{slr} over the 71 year design life of a hypothetical house, we seek to answer the research question ``\emph{how can deep uncertainties be transparently synthesized for decision analysis?}''

@@

Following \cref{fig:flowchart}, we consider using $J$ \glspl{sow}, $\vb{s} = \qty{s_1, s_2, \ldots, s_J}$, to evaluate $I$ candidate decisions, $\vb{x} = \qty{x_1, x_2, \ldots, x_I}$.
For each scenario $s_j \in \vb{s}$ and decision $x_i \in \vb{x}$ we use a system model $f$ to calculate a set of metrics describing the performance of decision $x_i$ on \gls{sow} $s_j$, which we denote $u_{ij} = f(x_i, s_j)$.
While we assume for simplicity that the decision space is known and finite, this approach could be coupled to a policy search model that proposes candidate decisions.

@@

A first analytical step is to use the model in an ``exploratory'' mode.
Exploratory modeling is averse to making explicit assumptions about the likelihood of different \glspl{sow} and instead seek to generate new knowledge \citep{bankes:1993}by systematically exploring a large number of possible futures, emphasizing interactions between different uncertainties \citep{reed_msdbook:2022}.
Exploratory modeling is often paired with analyses that identify relevant scenarios  \citep{lamontagne_discovery:2018,groves_scenarios:2007} or summarize a system's response to forcing \citep{Poff:2015jn,Steinschneider:2015kk,sriver_sealevel:2018}.
Despite the aversion to strong assumptions about the likelihood of different futures, subjective modeling decisions such as the choice of system modle, the set of candidate decisions, the criteria used to assess outcomes, and the choice of how to sample \glspl{sow} strongly influence results \citep{quinn_exploratory:2020}.

@@

Although exploratory modeling is a useful framework for understanding systems, there are many questions that it cannot answer.
For example, answering questions like ``what is the 95th percentile of metric $u$ under decision $x$'' or ``which decision minimizes expected damages'' or ``what is the probability of exceeding a critical threshold'' requires constructing estimators for the probability distribution over outcomes.\klaus{Vivek suggests citing Jaynes here, anything in particular?}

@@

We can also use this approach to understand methods for \gls{dmdu} that sample a set of parameters from fixed ranges.
For example, \citet{sriver_sealevel:2018} sample parameters describing the rate of \gls{slr} across a range of values to inform coastal adaptation.
Similarly, \citet{trindade_waterpathways:2020} checks the performance of candidate decisions against an ensemble of synthetic time series of streamflow, water demand, and other parameters by sampling parameters that transform the available data over a plausible range \citep[\ie, robustness metrics; see][for details]{mcphail_robustness:2019,herman:2015}.
Since sampling over a range is equivalent to sampling from a Uniform distribution, this assumption is equivalent to assuming the true inputs to come from independent Uniform distributions (one for each parameter) bounded by the plausible range.

Our primary concern is not that subjective assumptions about the likelihood of different futures are wrong -- this is, almost surely, inevitable -- but that when these assumptions are opaque and presented without critique or validation they may lead to poor decisions.\james{Build this out -- it's not so much the potential for poor decisions so much as that it can be hard to undwind the decision and understand what's driving it. We cant know if critical assumptions turned out to be poor if we don't know what they are or how changing them would result in a different analysis}

@@

If the ensemble of \glspl{sow} is not drawn \gls{iid} from the true distribution (\ie, if we deliberately over-sample low-probability but high-impact \glspl{sow}) or if we wish to synthesize \emph{across} multiple scenarios $M_1, M_2, \ldots, M_K$, then additional measures must be taken.
One possibility is to discard the \glspl{sow} ($\vb{s}$) and resample from some ``true'' distribution, which we shall denote $p_\mathrm{belief}$.
In practice, this is impractical because simulating $\vb{s}$ may rely on complicated models that are computationally expensive to re-run.

Instead of attempting to re-sample $\vb{s}$, we instead consider re-weighting.
A simple approach would be to assign each \gls{sow} $s_j$ a weight $w_j$ equivalent to the ratio of the desired (often called target distribution in survey analysis) distribution, $p_\mathrm{belief}$ and the sampling distribution, which could be approximated through kernel density estimation or other methods.
Instead, paralleling joint probability methods for statistical analysis of tropical cyclones, we use a grid-based approach \citep{johnson_clara:2013,resio_probabilities:2007,toro_jpm-os:2010}.

First, we project the \glspl{sow} $\vb{s} \in \Omega$ onto a low-dimensional representation, which we denote $\qty{\psi_1, \psi_2, \ldots, \psi_J} \in \Psi$.
This allows us to specify a prior belief over this reduced space $p_\mathrm{belief}(\Psi)$ instead of over the full space $p_\mathrm{belief}(\Omega)$.
We then calculate a probabilistic weight $w_j \in \qty[0, 1]$ for each \gls{sow} $s_j$ so that the weighted distribution of $\psi_j$ closely approximates $p_\mathrm{belief}(\Psi)$.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{grid-sketch}
    \caption{
        Schematic of \gls{sow} weighting scheme defined in \cref{eq:weights}.
        This method is illustrated for a hypothetical target distribution (orange line) and $J=4$ samples $\psi_1, \psi_2, \psi_3, \psi_4$ (orange dots).
        As shown in \cref{eq:weights}, the weights $w_j$ (vertical lines) are calculated based on the cumulative distribution function of the target distribution  (black line) at the halfway points $\frac{1}{2}\qty[\psi_i+\psi_{i+1}]$ (vertical dashed lines).
    }\label{fig:grid-sketch}
\end{figure}

@@

\section{Case study}\label{sec:case-study}

To illustrate the proposed decision analytic framework we model a one-time decision of whether to elevate a house, and if so by how much (\cref{fig:xlrm}).
We focus on a case study of a \emph{hypothetical} house in Norfolk, VA.
For interpretability, we focus on deep uncertainty in \gls{msl} and treat other model parameters as more shallow uncertainties as shown in \cref{tab:uncertainties}.
We use the notation developed in the previous section to describe the case study.
Specifically,
\begin{enumerate}
    \item The decision vector $\vb{x}$ is comprised of discrete possible house heightenings ($\Delta h$); we consider $\Delta h = \qty{\SI{0}{ft}, \SI{0.25}{ft}, \ldots, \SI{12}{ft}}$ f
    \item The \glspl{sow} describe annual time series of \gls{msl} over the $T=71$ year house lifetime so $\vb{s} \in \mathbb{R}^T$
    \item The system model $f$ quantifies up-front costs and lifetime expected damages, given a decision $x_i$ and \gls{sow} $s_j$, by integrating economic and engineering damage models over a probability distribution for storm surge.
\end{enumerate}
In the remainder of this section we describe data sources and treatment of \gls{slr} (\cref{sec:case-slr}), storm surge (\cref{sec:case-surge}), damages and metrics (\cref{sec:case-metrics}), and finally the subjective priors $p_\mathrm{belief}$ used to apply the re-weighting method described in \cref{sec:analysis-synthesize} to this case study (\cref{sec:case-priors}).

\begin{table}
    \centering
    \caption{
        Summary of parameters, their notation, and how their uncertainty is represented.
        Symbols describing the decision-analytic framework are described in \cref{fig:flowchart}.
    }\label{tab:uncertainties}
    \footnotesize
    \begin{tabular}{p{1.25in} p{0.75in} p{3in}}
        \toprule
        Name                 & Symbol            & Uncertainty                                                                          \\
        \midrule
        \Gls{msl}            & $\overline{y}(t)$ & Deeply uncertain: four physical models $\times$ four \acrshort{rcp} scenarios        \\
        Storm surge          & $y'(t)$           & Probabilistic: Bayesian inference on a stationary \acrshort{gev} model               \\
        Annual maximum flood & $y(t)$            & Deterministic: $y(t)=\overline{y}(t)+y'(t)$                                          \\
        Discount rate        & $1-\gamma$        & Fixed at 2.5\% per year                                                              \\
        Depth-damage         & $D(h-y)$          & Deterministic: based on HAZUS model \citep[see][]{zarekarizi_suboptimal:2020}        \\
        Elevation cost       & $C(\Delta h)$     & Deterministic: a piecewise linear model following \citet{zarekarizi_suboptimal:2020} \\
        Initial height       & $h_0$             & Deterministic: \SI{1}{ft} below the \gls{bfe}, unless otherwise noted                \\
        House floor area     & --                & Deterministic: \SI{1500}{ft^2}                                                       \\
        Structural value     & --                & Deterministic: \usd{200000}                                                          \\
        House lifespan       & $T$               & Deterministic: 71 years                                                              \\
        \bottomrule
    \end{tabular}
    %TODO: check if there are other symbols used
\end{table}

@@

Data on storm surge comes from Sewells Point, VA (gauge 8638610) from the NOAA tides and currents dataset,  freely available to the public at \url{https://tidesandcurrents.noaa.gov/waterlevels.html}.\klaus{I can't find a better reference for this dataset!}
Hourly recordings of water level are available from 1928 to the present; we use data from the period January 1, 1928 to December 31, 2021.
For each calendar year we first remove the annual mean, then calculate the maximum water level.
We refer this time series of annual maximum storm surges as $y'(t)$.
We display this time series of annual maxima storm surges in \cref{fig:surge-obs-return}(a).
The largest recorded surge was the Chesapeake-Potomac hurricane of 1933, which caused a surge of over \SI{7}{ft} at this gauge, but other hurricanes and Nor'easters have caused surges above \SI{6}{ft}.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{surge-obs-return}
    \caption{
        Annual maximum storm surges (after subtracting mean sea level) at Sewells Point, VA.
        (A):
        time series of historic storms.
        Red (yellow) arrows denote notable tropical cyclones (Nor'easters).
        (B):
        return periods.
        Dots indicate observed values; their $x$-value is their plotting position using the Weibull formula (eq.~\ref{eq:weibull}).
        Gray lines show the 50, 80, and 95\% posterior confidence intervals from the Bayesian \gls{gev} fit (\cref{sec:case-surge}).
    }\label{fig:surge-obs-return}
\end{figure}

We model future storm surge using a stationary \gls{gev} model:
\begin{equation}\label{eq:surge-model}
    y'(t) \sim \text{GEV}\qty(\mu, \sigma, \xi),
\end{equation}
where $y'(t)$ is the storm surge (above sea level) in year $t$ and a \gls{gev} distribution with location, shape, and scale parameters $\mu$, $\sigma$, and $\xi$, respectively, has the probability density function given in \cref{eq:gev-dist}.
This model assumes stationarity, neglecting any potential time dependence.

>>>>

% Figures
\usepackage{graphicx}
\usepackage[list=true]{subcaption}
\graphicspath{{../../plots/}{../../tikz/}{../../img/}}
\usepackage[section]{placeins} % require floats to appear in the section they are defined

@@

% ACRONYMS
\usepackage[acronym, nopostdot, nonumberlist, shortcuts, numberedsection, nogroupskip,]{glossaries}
\newacronym{bfe}{BFE}{base flood elevation}
\newacronym{cmip}{CMIP}{the Coupled Model Intercomparison Project}
\newacronym{dmdu}{DMDU}{decision making under deep uncertainty}
\newacronym{fema}{FEMA}{the Federal Emergency Management Agency}
\newacronym{gev}{GEV}{generalized extreme value}
\newacronym{hazus}{HAZUS}{Hazard U.S.}
\newacronym{iid}{IID}{independent and identically distributed}
\newacronym{ipcc}{IPCC}{International Panel on Climate Change}
\newacronym{msl}{MSL}{mean relative sea level}
\newacronym{noaa}{NOAA}{the National Oceanic and Atmospheric Administration}
\newacronym{pdf}{PDF}{probability density function}
\newacronym{rcp}{RCP}{representative concentration pathway}
\newacronym{slr}{SLR}{sea level rise}
\newacronym{ssp}{SSP}{shared socio-economic pathway}
\newacronym[]{usace}{USACE}{United States Army Corps of Engineers}
\newacronym[]{usgs}{USGS}{United States Geological Survey}
\newacronym[plural=SOWs,firstplural=states of the world (SOWs),descriptionplural=states of the world]{sow}{SOW}{state of the world}
\makeglossaries % required to make the list of acronyms

@@

Currently used methods for estimating these \glspl{pdf} emphasize data-driven, nominally objective methods that can be applied consistently across locations.
For example, \gls{usgs} Bulletin 17C specifies procedures for estimating flood frequency and \gls{noaa} Atlas 14 provides estimates of the intensity, duration, and frequency of extreme rainfall.
Among several statistical assumptions these analyses make that have been recently called into question is stationarity \citep[the assumption that past and future hazard come from the same \gls{pdf}; see][for a review]{merz_review:2014}.
For example, clear trends in extreme rainfall are apparent across much of southeastern Texas \citep{fagnant_spatiotemporal:2020,nielsen-gammon_txrainfall:2020}, and trends in many other hazards are consistent with observations \citep[see][for a comprehensive summary]{ipcc_impacts:2022}.\klaus{I'm citing the technical summary which is 96 pages -- there are about 100 authors listed. Any suggestion for how to cite best?}
While some methods have been proposed for incorporating trends into these analyses \citep[see][for a review]{Salas:2018ge}, these assume specific forms of nonstationarity which may not fully represent physical processes or true levels of uncertainty \citep{DossGollin:2019,Montanari:2014hl,Serinaldi:2015bq}.
Ultimately, the difficulty of incorporating nonstationarity into existing frameworks has led to continued reliance on stationarity despite recognition of the associated limitations (see \cite{bulletin17c:2019}, p.~2, or \cite{atlas14_texas:2018} p.~A.4-42).

@@

Although these methods for \gls{dmdu} have proven valuable in a wide range of settings, they still require necessarily subjective assessments about the likelihood of future conditions.
Even exploratory models require the analyst to choose which uncertainties are considered and how they are sampled.
When analyses use metrics that integrate performance over many possible futures, whether to compute sensitivities, estimate expectations, or compute robustness metrics, they necessarily make assumptions about the likelihood of different futures.
For example, uniform priors are often justified by deferring to Laplace's principle of insufficient reason \citep[see][p.~135]{stigler_uncertainty:1986}, but even this choice makes strong assumptions that are sensitive to factors such as parameterization \citep[p.54]{gelman_bda3:2014}.
While one interpretation of deep uncertainty is that that specifying a joint \gls{pdf} over inputs is inappropriate, assumptions about the ranges and independence of parameters to sample are just as subjective as the choice of probability distribution \citep{schneider_scenarios:2002,quinn_exploratory:2020} and, indeed, can be interpreted  as a specific choice of probability distribution (we revisit this point in \cref{sec:analysis-condition}).
This motivates the development of decision analytic frameworks that draw from the strengths of \gls{dmdu} methods such as exploratory modeling, vulnerability assessment, robustness checks, and iterative stakeholder critique, but that embrace the reality that assumptions about the future are inescapable.

@@

We are motivated by parallels between decision making under deep uncertainty and the statistical problem of model selection.
\citet{oreskes_verification:1994} argues that because natural systems are never closed and model results are never unique, validation and verification of models representing these systems is necessarily qualitative and subjective.
In the literature on decision making under deep uncertainty, this has led to recognition of the need to develop strategies that are robust to model imperfections.
We find parallels to this challenge in the Bayesian literature on model selection, particularly in the ``$\mathcal{M}$-open'' case in which there is no ``true'' model to identify and all models are imperfect representations of reality.
This is a theoretical contrast to the ``$\mathcal{M}$''-closed case in which a true model can be identified and written down but is one amongst finitely many models from which an analyst has to choose as well as the ``$\mathcal{M}$''-complete case in which a true model exists but is inaccessible in the sense that even though it can be conceptualized it cannot be written down \citep{bernardo_bayesian:1994}.
Whereas model selection in the $\mathcal{M}$-closed case seeks to asymptotically identify the true model from a set of candidate models, model selection in the $\mathcal{M}$-open case emphasizes a subjective view, both of probability itself and of the modeling process, for which probability offers a self-consistent language with which to reason about the unknown rather than a statement of objective truth (see \cite{gelman_philosophy:2013}, \cite{ramsey_probability:2016}, or \cite{jaynes_probability:2003} for a discussion of Bayesian philosophy and \cite{Piironen:2017eh} for a technical discussion of methods for model selection).
From a practical perspective, model selection in the $\mathcal{M}$-open case emphasizes building models iteratively, simulating the consequences of these models, and subsequently using these simulations to critique and improve them \citep{gelman_workflow:2020}.

In this paper we offer a first conceptual step towards bridging the fields of \gls{dmdu} and Bayesian model building.
We consider a didactic case study of whether to elevate a hypothetical house, and if so how high, as a specific example of a decision problem subject to both shallow (storm surge) and deep (\gls{slr}) uncertainties.
Prior studies have found that floodproofing and building-scale vulnerability reduction measures, including house elevation, can effectively reduce local flood damages in many contexts \citep{demoel_reducing:2014,deruig_building:2020,kreibich_building:2005,slotter_floodproofing:2020,rozer_coping:2016,mobley_mitigation:2020,aerts_cost:2018}, and both local building codes \citep{asce_7-10:2013,bruneau_multihazard:2017,asce_24-05:2006} and federal policy \citep{FEMA_p-55:2011} require elevation in some cases.
Guidance for homeowners, notably from \gls{fema}, recommends elevating to the \gls{bfe} (typically the \SI{100}{year} flood) plus a freeboard \citep{fema_retrofitting:2014,asce_24-14:2015,fema_retrofitting:2014} but recent research has demonstrated that neglecting uncertainty in the cost-benefit analysis can lead to poor decisions \citep{zarekarizi_suboptimal:2020}.
Focusing on deep uncertainty in \gls{slr} over the 70\james{Need to re-run codes with 70 rather than 71} year design life of a hypothetical house, we seek to answer the research question ``\emph{how can deep uncertainties be transparently synthesized for decision analysis?}''

@@

Following \cref{fig:flowchart}, we consider using $J$ \glspl{sow}, $\vb{s} = \qty{s_1, s_2, \ldots, s_J}$, to evaluate $I$ candidate decisions, $\vb{x} = \qty{x_1, x_2, \ldots, x_I}$.
For each scenario $s_j \in \vb{s}$ and decision $x_i \in \vb{x}$ we use a system model $f$ (comprised of multiple components) to calculate a set of metrics describing the performance of decision $x_i$ on \gls{sow} $s_j$, which we denote $u_{ij} = f(x_i, s_j)$.
While we assume for simplicity that the decision space is known and finite, this approach could be coupled to a policy search model that proposes candidate decisions.

@@

A first analytical step is to use the model in an ``exploratory'' mode.
Exploratory modeling is averse to making explicit assumptions about the likelihood of different \glspl{sow} and instead seek to generate new knowledge \citep{bankes:1993} by systematically exploring a large number of possible futures, emphasizing interactions between different uncertainties \citep{reed_msdbook:2022}.
Exploratory modeling is often paired with analyses that identify relevant scenarios  \citep{lamontagne_discovery:2018,groves_scenarios:2007} or summarize a system's response to forcing \citep{Poff:2015jn,Steinschneider:2015kk,sriver_sealevel:2018}.
Despite the aversion to strong assumptions about the likelihood of different futures, subjective modeling decisions such as the choice of system model, the set of candidate decisions, the criteria used to assess outcomes, and the choice of how to sample \glspl{sow} strongly influence results \citep{quinn_exploratory:2020}.

@@

Although exploratory modeling is a useful framework for understanding systems, there are many questions that it cannot answer.
For example, answering questions like ``what is the 95th percentile of metric $u$ under decision $x$'' or ``which decision minimizes expected damages'' or ``what is the probability of exceeding a critical threshold'' requires constructing estimators for the probability distribution over outcomes \citep{schneider_scenarios:2002}.

@@

We can also use this approach to understand \gls{dmdu} methods that sample a set of parameters from fixed ranges.
For example, \citet{sriver_sealevel:2018} sample parameters describing the rate of \gls{slr} across a range of values to inform coastal adaptation.
Similarly, \citet{trindade_waterpathways:2020} checks the performance of candidate decisions against an ensemble of synthetic time series of streamflow, water demand, and other parameters by sampling parameters that transform the available data over a plausible range \citep[\ie, robustness metrics; see][for details]{mcphail_robustness:2019,herman:2015}.
Since sampling over a range is equivalent to sampling from a Uniform distribution, this assumption is equivalent to assuming the true inputs to come from independent Uniform distributions (one for each parameter) bounded by the plausible range.

Our primary concern is not that subjective assumptions about the likelihood of different futures are wrong -- this is, almost surely, inevitable -- but that when these assumptions are opaque and presented without critique or validation they may lead to poor decisions.
\james{Build this out -- it's not so much the potential for poor decisions so much as that it can be hard to undwind the decision and understand what's driving it. We cant know if critical assumptions turned out to be poor if we don't know what they are or how changing them would result in a different analysis. That said, I think they can also lead to poor decisions!}

@@

If the \gls{sow} ensemble is not drawn \gls{iid} from the true distribution (\eg, if one deliberately over-samples low-probability but high-impact \glspl{sow}) or if one wishes to synthesize \emph{across} multiple scenarios $M_1, M_2, \ldots, M_K$, then additional measures must be taken.
One possibility is to discard the \glspl{sow} ($\vb{s}$) and resample from some ``true'' distribution, which we shall denote $p_\mathrm{belief}$.
In practice, this is impractical because simulating $\vb{s}$ may rely on complicated models that are computationally expensive to re-run.

Instead of attempting to re-sample $\vb{s}$, we instead consider re-weighting.
A simple approach would be to assign each \gls{sow} $s_j$ a weight $w_j$ equivalent to the ratio of the desired (often called target distribution in survey analysis) distribution, $p_\mathrm{belief}$ and the sampling distribution, which could be approximated through methods such as kernel density estimation.
Paralleling joint probability methods for statistical analysis of tropical cyclones, we use a grid-based approach \citep{johnson_clara:2013,resio_probabilities:2007,toro_jpm-os:2010}.
First, we project the \glspl{sow} $\vb{s} \in \Omega$ onto a low-dimensional representation, which we denote $\qty{\psi_1, \psi_2, \ldots, \psi_J} \in \Psi$.
This allows us to specify a prior belief over this reduced space $p_\mathrm{belief}(\Psi)$ instead of over the full space $p_\mathrm{belief}(\Omega)$.
We then calculate a probabilistic weight $w_j \in \qty[0, 1]$ for each \gls{sow} $s_j$ so that the weighted distribution of $\psi_j$ closely approximates $p_\mathrm{belief}(\Psi)$.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{grid-sketch}
    \caption{
        Schematic of \gls{sow} weighting scheme defined in \cref{eq:weights}.
        This method is illustrated for a hypothetical target distribution (orange line) and $J=4$ samples $\psi_1, \psi_2, \psi_3, \psi_4$ (orange dots).
        As shown in \cref{eq:weights}, the weights $w_j$ (vertical lines) are calculated based on the cumulative distribution function of the target distribution  (black line) at the halfway points $\frac{1}{2}\qty[\psi_i+\psi_{i+1}]$ (vertical dashed lines).
        % TODO: add the weights to the legend
        % TODO: clarify colors! (NO, Klaus, you're not color blind -- but I am a little bit)
    }\label{fig:grid-sketch}
\end{figure}

@@

\section{Demonstrating the concept with a case study}\label{sec:case-study}

To illustrate the proposed decision analytic framework, we model a one-time decision of whether to elevate a house, and if so by how much (\cref{fig:xlrm}).
We focus on a case study of a \emph{hypothetical} house in Norfolk, VA.
For interpretability, we focus on deep uncertainty in \gls{msl} and approximate other model parameters as shallow uncertainties as shown in \cref{tab:uncertainties}.
We use the notation developed in the previous section to describe the case study.
Specifically,
\begin{enumerate}
    \item The decision vector $\vb{x}$ is comprised of discrete possible house heightenings ($\Delta h$); we consider $\Delta h = \qty{\SI{0}{ft}, \SI{0.25}{ft}, \ldots, \SI{12}{ft}}$ f
    \item The \glspl{sow} describe annual time series of \gls{msl} over the $T=70$ year house lifetime so $\vb{s} \in \mathbb{R}^T$
    \item The system model $f$ quantifies up-front costs and lifetime expected damages, given a decision $x_i$ and \gls{sow} $s_j$, by integrating economic and engineering damage models over a probability distribution for storm surge.
\end{enumerate}
In the remainder of this section we describe data sources and treatment of \gls{slr} (\cref{sec:case-slr}), storm surge (\cref{sec:case-surge}), damages and metrics (\cref{sec:case-metrics}), and finally the subjective priors $p_\mathrm{belief}$ used to apply the re-weighting method described in \cref{sec:analysis-synthesize} to this case study (\cref{sec:case-priors}).

\begin{table}
    \centering
    \caption{
        Summary of parameters, their notation, and how their uncertainty is represented.
        Symbols describing the decision-analytic framework are described in \cref{fig:flowchart}.
    }\label{tab:uncertainties}
    \footnotesize
    \begin{tabular}{p{1.25in} p{0.75in} p{3in}}
        \toprule
        Name                 & Symbol            & Uncertainty                                                                          \\
        \midrule
        \Gls{msl}            & $\overline{y}(t)$ & Deeply uncertain: four physical models $\times$ four \acrshort{rcp} scenarios        \\
        Storm surge          & $y'(t)$           & Probabilistic: Bayesian inference on a stationary \acrshort{gev} model               \\
        Annual maximum flood & $y(t)$            & Deterministic: $y(t)=\overline{y}(t)+y'(t)$                                          \\
        Discount rate        & $\rho$            & Determinitic: 2.5\% per year                                                         \\
        Depth-damage         & $D(h-y)$          & Deterministic: based on HAZUS model \citep[see][]{zarekarizi_suboptimal:2020}        \\
        Elevation cost       & $C(\Delta h)$     & Deterministic: a piecewise linear model following \citet{zarekarizi_suboptimal:2020} \\
        Initial height       & $h_0$             & Deterministic: \SI{1}{ft} below the \gls{bfe}, unless otherwise noted                \\
        House floor area     & --                & Deterministic: \SI{1500}{ft^2}                                                       \\
        Structural value     & --                & Deterministic: \usd{200000}                                                          \\
        House lifespan       & $T$               & Deterministic: 70 years                                                              \\
        \bottomrule
    \end{tabular}
    %TODO: check if there are other symbols used
\end{table}

@@

Data on storm surge comes from Sewells Point, VA (gauge 8638610) from the NOAA tides and currents dataset,  freely available to the public at \url{https://tidesandcurrents.noaa.gov/waterlevels.html}.
Hourly recordings of water level are available from 1928 to the present; we use data from the period January 1, 1928 to December 31, 2021.
For each calendar year we first remove the annual mean, then calculate the maximum water level.
We refer this time series of annual maximum storm surges as $y'(t)$.
We display this time series of annual maxima storm surges in \cref{fig:surge-obs-return}(a).
The largest recorded surge was the Chesapeake-Potomac hurricane of 1933, which caused a surge of over \SI{7}{ft} at this gauge, but other hurricanes and Nor'easters have caused surges above \SI{6}{ft}.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{surge-obs-return}
    \caption{
        Annual maximum storm surges (after subtracting \acrlong{msl}) at Sewells Point, VA from NOAA tides and currents dataset available at \url{https://tidesandcurrents.noaa.gov/waterlevels.html}.
        (A):
        time series of historic storms.
        Red (yellow) arrows denote notable tropical cyclones (Nor'easters).
        (B):
        return periods.
        Dots indicate observed values; their $x$-value is their plotting position using the Weibull formula (eq.~\ref{eq:weibull}).
        Gray lines show the 50, 80, and 95\% posterior confidence intervals from the Bayesian \gls{gev} fit (\cref{sec:case-surge}).
        % TODO: change caption for Obs (Weibull Plot Pos.) => Observations
        % TODO: make the different confidence intervals slightly different colors instead of changing opacity
        % TODO: move (B) to more logical location
    }\label{fig:surge-obs-return}
\end{figure}

We model future storm surge using a stationary \gls{gev} model:
\begin{equation}\label{eq:surge-model}
    y'(t) \sim \text{GEV}\qty(\mu, \sigma, \xi),
\end{equation}
where $y'(t)$ is the storm surge (above \gls{msl}) in year $t$ and a \gls{gev} distribution with location, shape, and scale parameters $\mu$, $\sigma$, and $\xi$, respectively, has the probability density function given in \cref{eq:gev-dist}.
This model assumes stationarity, neglecting any potential time dependence.
